---
title: "Preterm Birth Prediction: Clinical & Environmental Features"
subtitle: "End-to-end analysis: cleaning, EDA, GLM, and Random Forest"
author: "Hanyu Song"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 8,
  fig.height = 6,
  dpi        = 150
)
```

# Setup

```{r load-packages}
suppressPackageStartupMessages({
  # data wrangling & I/O
  library(tidyverse)
  library(janitor)
  library(skimr)
  # tables & summaries
  library(gtsummary)
  library(gt)
  # plots
  library(ggcorrplot)
  library(maps)
  # modeling
  library(ranger)
  library(rsample)
  library(pROC)
  library(ResourceSelection)
  library(lmtest)
})

here::i_am("codes/full_analysis.Rmd")
```

```{r ensure-output-dirs, include=FALSE}
# Make sure expected sub-directories exist (in case this is the first run)
for (d in c("outputs", "figures", "tables")) {
  dir_path <- here::here(d)
  if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
}
```

---

# Data Cleaning

## Load raw data

```{r load-raw}
t1_raw <- suppressMessages(read_csv(here::here("data", "EPIC_GA_status.csv")))
t2_raw <- suppressMessages(read_csv(here::here("data", "pull.v1.csv")))
```

## Basic cleaning & merge

```{r basic-clean}
# 1.1 Cleaning t1_raw
t1 <- t1_raw %>%
  clean_names() %>%
  mutate(encounter_nbr = as.character(encounter_nbr)) %>%
  select(encounter_nbr, admission_date, zip_code, lat, long, score, status)

# 1.2 Cleaning t2_raw
t2 <- t2_raw %>%
  clean_names() %>%
  mutate(encounter_nbr = as.character(encounter_nbr)) %>%
  select(-x1, -address, -city, -state, -zip_code)

# 1.3 Merge t1 and t2
data_clean <- t2 %>%
  left_join(t1, by = "encounter_nbr")

data_clean <- data_clean %>%
  mutate(
    across(where(is.character), as.factor),
    across(c(preterm, obesity, gestht, htn, prom, gestdiab,
             preecl_s, fgr, diab, preecl_n, nulliparous), as.factor),
    # set white as reference group
    race = relevel(race, ref = "White")
  )
```

## Missing values summary

```{r missing-summary}
missing_summary <- data_clean %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to  = "Variable",
               values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = (Missing_Count / nrow(data_clean)) * 100) %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

missing_summary
```

## Advanced cleaning

```{r advanced-clean}
# 2.1 Remove missing values and duplicate encounter_nbr
data_clean <- data_clean %>%
  drop_na() %>%
  distinct(encounter_nbr, .keep_all = TRUE)

# 2.2 Remove the outlier in birth_weight_oz
data_clean <- data_clean %>%
  filter(encounter_nbr != "127788738")

# 2.3 Remove the outlier found in group preterm = 0
outlier_check <- data_clean %>%
  filter(preterm == "0") %>%
  arrange(birth_weight_oz) %>%
  select(encounter_nbr, gest_age, birth_weight_oz) %>%
  head(10)

outlier_check

data_final_cleaned <- data_clean %>%
  filter(encounter_nbr != "149794760")
```

## Save cleaned data

```{r save-cleaned}
saveRDS(data_final_cleaned, here::here("outputs", "01_dataclean.Rds"))
write.csv(data_final_cleaned, here::here("outputs", "01_dataclean.csv"),
          row.names = FALSE)
```

---

# Exploratory Analysis

## Table 1 — Stratified by race

```{r table1, results='asis'}
table1 <- data_final_cleaned %>%
  select(
    # Demographics
    race, age, age_cat, hisp, hospital, nulliparous, score,
    # Clinical risk factors
    obesity, htn, gestht, diab, gestdiab, preecl_n, preecl_s, prom, fgr,
    # Outcomes
    gest_age, birth_weight_oz, preterm
  ) %>%
  tbl_summary(
    by = race,
    label = list(
      age            ~ "Maternal Age (years)",
      age_cat        ~ "Age Category",
      hisp           ~ "Hispanic Ethnicity",
      hospital       ~ "Hospital Location",
      nulliparous    ~ "Nulliparous (First birth)",
      obesity        ~ "Obesity (BMI >= 30)",
      htn            ~ "Chronic Hypertension",
      gestht         ~ "Gestational Hypertension",
      diab           ~ "Pre-gestational Diabetes",
      gestdiab       ~ "Gestational Diabetes",
      preecl_n       ~ "Preeclampsia (Non-severe)",
      preecl_s       ~ "Severe Preeclampsia",
      prom           ~ "Premature Rupture of Membranes",
      fgr            ~ "Fetal Growth Restriction",
      gest_age       ~ "Gestational Age at Delivery (weeks)",
      birth_weight_oz ~ "Birth Weight (oz)",
      preterm        ~ "Preterm Birth (<37 weeks)"
    ),
    statistic = list(all_continuous()  ~ "{mean} ({sd})",
                     all_categorical() ~ "{n} ({p}%)"),
    missing = "no",
    digits  = all_continuous() ~ 1
  ) %>%
  add_p(
    test      = list(all_categorical() ~ "chisq.test"),
    test.args = list(all_categorical() ~ list(simulate.p.value = TRUE))
  ) %>%
  add_overall() %>%
  bold_labels()

table1

# Save the gt version to disk
table1 %>%
  as_gt() %>%
  gt::gtsave(here::here("tables", "02_table1.html"))
```

## Preterm birth rate by race

```{r preterm-rate-plot, fig.width=12, fig.height=5}
plot_data <- data_final_cleaned %>%
  group_by(race) %>%
  summarise(
    total         = n(),
    preterm_count = sum(preterm == "1"),
    preterm_rate  = preterm_count / total
  )

preterm_birth_rate_plot <- ggplot(
  plot_data,
  aes(x = reorder(race, -preterm_rate), y = preterm_rate, fill = race)
) +
  geom_col() +
  geom_text(aes(label = scales::percent(preterm_rate, accuracy = 0.1)),
            vjust = -0.5) +
  labs(title = "Preterm Birth Rates by Race",
       y = "Preterm Rate", x = "Race") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "none", panel.grid = element_blank())

preterm_birth_rate_plot

ggsave(
  filename = here::here("figures", "02_preterm_rate_plot.png"),
  plot     = preterm_birth_rate_plot,
  width    = 15, height = 6, dpi = 300
)
```

## Birth weight by preterm status

```{r birth-weight-dist}
birth_weight_dist_plot <- ggplot(
  data_final_cleaned,
  aes(x = preterm, y = birth_weight_oz, fill = preterm)
) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Birth Weight Distribution by Preterm Status",
       x = "Preterm (0 = No, 1 = Yes)",
       y = "Birth Weight (oz)") +
  theme_classic()

birth_weight_dist_plot

ggsave(
  filename = here::here("figures", "02_birth_weight_dist_plot.png"),
  plot     = birth_weight_dist_plot,
  width    = 8, height = 6, dpi = 300
)
```

## Correlation of clinical risk factors

```{r correlation-heatmap}
cor_data <- data_final_cleaned %>%
  select(obesity, htn, gestht, diab, gestdiab,
         preecl_n, preecl_s, prom, fgr) %>%
  mutate(across(everything(), ~ as.numeric(as.character(.))))

corr_matrix <- cor(cor_data)

clinical_rf_heatmap <- ggcorrplot(
  corr_matrix,
  method = "square", type = "lower", lab = TRUE,
  title  = "Correlation Matrix of Risk Factors"
) +
  theme_minimal() +
  theme(panel.grid = element_blank())

clinical_rf_heatmap

ggsave(
  filename = here::here("figures", "02_clinical_rf_heatmap.png"),
  plot     = clinical_rf_heatmap,
  width    = 8, height = 6, dpi = 300
)
```

## Spatial distribution of births

```{r spatial-map, fig.width=6, fig.height=8}
ga_counties <- map_data("county", region = "georgia")

spatial_dis_map <- ggplot() +
  geom_polygon(
    data = ga_counties,
    aes(x = long, y = lat, group = group),
    fill = "white", color = "black", linewidth = 0.2
  ) +
  geom_point(
    data = data_final_cleaned,
    aes(x = long, y = lat, color = preterm),
    alpha = 0.5, size = 0.5
  ) +
  labs(title    = "Spatial Distribution of Births",
       subtitle = "Red points indicate preterm births",
       color    = "Preterm") +
  coord_fixed(1.3) +
  theme_void() +
  scale_color_manual(values = c("grey80", "red"))

spatial_dis_map

ggsave(
  filename = here::here("figures", "02_spatial_dis_map.png"),
  plot     = spatial_dis_map,
  width    = 6, height = 8, dpi = 300
)
```

---

# Logistic Regression (GLM)

## Merge clinical data with AlphaEarth feature `A59`

```{r merge-alphaearth-glm}
alphaearth_data <- read_csv(
  here::here("outputs", "Capstone_AlphaEarth_2023_1000m_extracted.csv"),
  show_col_types = FALSE
) %>%
  mutate(encounter_nbr = as.character(encounter_nbr)) %>%
  select(encounter_nbr, A59)

data_all_cleaned <- data_final_cleaned %>%
  mutate(encounter_nbr = as.character(encounter_nbr)) %>%
  left_join(alphaearth_data, by = "encounter_nbr")

# set the reference race to white (already done in cleaning, but keep explicit)
data_all_cleaned$race <- relevel(as.factor(data_all_cleaned$race), ref = "White")
```

## Base clinical model

```{r glm-base}
base_model <- glm(
  preterm ~ preecl_s + age + I(age^2) + fgr + prom + htn + preecl_n +
    race + diab + hospital + hisp + nulliparous + gestdiab + obesity + gestht,
  data   = data_all_cleaned,
  family = binomial(link = "logit")
)

summary(base_model)
```

## Clinical model with `htn × obesity` interaction

```{r glm-interaction}
interaction_model <- glm(
  preterm ~ preecl_s + age + I(age^2) + fgr + prom + htn + preecl_n +
    race + diab + hospital + hisp + nulliparous + gestdiab + obesity + gestht +
    htn:obesity,
  data   = data_all_cleaned,
  family = binomial(link = "logit")
)

summary(interaction_model)
```

## Clinical + best spatial feature (`A59`) model

```{r glm-spatial}
spatial_model <- glm(
  preterm ~ preecl_s + age + I(age^2) + fgr + prom + htn + preecl_n +
    race + diab + hospital + hisp + nulliparous + gestdiab + obesity + gestht +
    A59,
  data   = data_all_cleaned,
  family = binomial(link = "logit")
)

summary(spatial_model)
```

## Likelihood-ratio comparisons

```{r glm-lrtests}
# Compare clinical vs clinical + interaction
lrtest(base_model, interaction_model)

# Compare clinical vs clinical + best spatial feature
lrtest(base_model, spatial_model)
```

## Table 2 — final spatial model

```{r glm-table2}
tbl_model <- spatial_model %>%
  tbl_regression(exponentiate = TRUE) %>%
  bold_p() %>%
  add_global_p() %>%
  bold_labels()

tbl_model

tbl_model %>%
  as_gt() %>%
  gt::gtsave(here::here("tables", "03_table2_model.html"))
```

## Goodness of fit (Hosmer–Lemeshow)

```{r glm-hoslem}
hoslem.test(spatial_model$y, fitted(spatial_model))
```

---

# Random Forest

## Build modeling dataset

```{r rf-build-data}
all_feature_data <- read_csv(
  here::here("outputs", "Capstone_AlphaEarth_2023_1000m_extracted.csv"),
  show_col_types = FALSE
) %>%
  mutate(encounter_nbr = as.character(encounter_nbr))

all_clinical_data <- read_csv(
  here::here("outputs", "01_dataclean.csv"),
  show_col_types = FALSE
) %>%
  mutate(encounter_nbr = as.character(encounter_nbr))

combined_data <- all_clinical_data %>%
  inner_join(all_feature_data, by = "encounter_nbr") %>%
  select(-any_of(c(
    "...1", "encounter_nbr",
    "gest_age", "age_cat",
    "birth_weight_oz",
    "lat", "long", ".geo",
    "system:index",
    "zip_code", "score",
    "status",
    "admission_date.x", "admission_date.y",
    "buffer_m", "center_lat", "center_long"
  ))) %>%
  mutate(
    preterm = as.factor(preterm),
    across(where(is.character), as.factor)
  )

dim(combined_data)
head(combined_data)
```

## Train / test split (stratified on `preterm`)

```{r rf-split}
set.seed(2026)
data_split <- initial_split(combined_data, prop = 0.8, strata = preterm)

train_data <- training(data_split)
test_data  <- testing(data_split)

dim(train_data)
dim(test_data)

prop.table(table(train_data$preterm))
prop.table(table(test_data$preterm))
```

## Full model (all variables)

```{r rf-full, cache=TRUE}
n_tree <- 500

rf_model <- ranger(
  formula     = preterm ~ .,
  data        = train_data,
  num.trees   = n_tree,
  importance  = "impurity",
  probability = TRUE,
  seed        = 2026
)

print(rf_model)
saveRDS(rf_model,
        here::here("outputs", paste0("03_rf_full_model_", n_tree, "trees.Rds")))
```

## Clinical-only model

```{r rf-clinical, cache=TRUE}
train_data_clinical_only <- train_data %>%
  select(-starts_with("A", ignore.case = FALSE))
test_data_clinical_only  <- test_data %>%
  select(-starts_with("A", ignore.case = FALSE))

dim(train_data_clinical_only)
dim(test_data_clinical_only)

rf_model_clinical <- ranger(
  formula     = preterm ~ .,
  data        = train_data_clinical_only,
  num.trees   = n_tree,
  importance  = "impurity",
  probability = TRUE,
  seed        = 2026
)

print(rf_model_clinical)
saveRDS(rf_model_clinical,
        here::here("outputs", paste0("03_rf_clinical_model_", n_tree, "trees.Rds")))

# Baseline ROC for DeLong tests
pred_probs_clinical <- predict(rf_model_clinical, data = test_data_clinical_only)$predictions[, "1"]
rf_roc_clinical <- roc(test_data_clinical_only$preterm, pred_probs_clinical,
                       levels = c("0", "1"), quiet = TRUE)
```

## Search for best single environmental feature

```{r rf-search-best-feature, cache=TRUE}
spatial_features <- sprintf("A%02d", 0:63)

auc_delong_results <- data.frame(
  Feature          = character(),
  AUC              = numeric(),
  DeLong_Pvalue    = numeric(),
  stringsAsFactors = FALSE
)

for (feat in spatial_features) {
  if (feat %in% names(train_data)) {

    temp_train <- train_data %>%
      select(-starts_with("A", ignore.case = FALSE), all_of(feat))
    temp_test  <- test_data %>%
      select(-starts_with("A", ignore.case = FALSE), all_of(feat))

    temp_rf <- ranger(
      formula     = preterm ~ .,
      data        = temp_train,
      num.trees   = n_tree,
      importance  = "none",
      probability = TRUE,
      seed        = 2026
    )

    temp_preds  <- predict(temp_rf, data = temp_test)$predictions[, "1"]
    temp_roc    <- roc(test_data$preterm, temp_preds,
                       levels = c("0", "1"), quiet = TRUE)
    temp_auc    <- as.numeric(temp_roc$auc)
    delong_test <- roc.test(rf_roc_clinical, temp_roc, method = "delong")

    auc_delong_results <- rbind(auc_delong_results, data.frame(
      Feature       = feat,
      AUC           = temp_auc,
      DeLong_Pvalue = delong_test$p.value
    ))
  }
}

final_evaluation_board <- auc_delong_results %>%
  arrange(desc(AUC)) %>%
  mutate(
    AUC           = round(AUC, 4),
    DeLong_Pvalue = round(DeLong_Pvalue, 4),
    Significant   = ifelse(DeLong_Pvalue < 0.05, "Yes", "No")
  )

head(final_evaluation_board, 15)

write.csv(final_evaluation_board,
          here::here("outputs", "03_All_Features_DeLong_Results.csv"),
          row.names = FALSE)
```

## Train RF with clinical + best feature

```{r rf-selected, cache=TRUE}
best_feature <- final_evaluation_board$Feature[1]
message("Best feature: ", best_feature,
        " (AUC = ", round(final_evaluation_board$AUC[1], 4), ")")

train_data_selected <- train_data %>%
  select(-starts_with("A", ignore.case = FALSE), all_of(best_feature))
test_data_selected  <- test_data %>%
  select(-starts_with("A", ignore.case = FALSE), all_of(best_feature))

rf_model_selected <- ranger(
  formula     = preterm ~ .,
  data        = train_data_selected,
  num.trees   = n_tree,
  importance  = "impurity",
  probability = TRUE,
  seed        = 2026
)

saveRDS(rf_model_selected,
        here::here("outputs", paste0("03_rf_selected_model_", n_tree, "trees.Rds")))
```

## ROC curve comparison

```{r rf-roc-plot, fig.width=8, fig.height=6}
# Full model
preds_probs_full <- predict(rf_model, data = test_data)$predictions[, "1"]
rf_roc_full <- roc(test_data$preterm, preds_probs_full,
                   levels = c("0", "1"), quiet = TRUE)

# Clinical + top environmental feature
pred_probs_selected <- predict(rf_model_selected, data = test_data_selected)$predictions[, "1"]
rf_roc_selected <- roc(test_data_selected$preterm, pred_probs_selected,
                       levels = c("0", "1"), quiet = TRUE)

plot(rf_roc_full,     col = "blue",  lwd = 2,
     main = "ROC Curve Comparison (Preterm Prediction)")
plot(rf_roc_clinical, col = "red",   lwd = 2, lty = 2, add = TRUE)
plot(rf_roc_selected, col = "green", lwd = 2, lty = 2, add = TRUE)

legend(
  "bottomright",
  legend = c(paste("Full Model (AUC:",     round(rf_roc_full$auc, 3),     ")"),
             paste("Clinical Model (AUC:", round(rf_roc_clinical$auc, 3), ")"),
             paste("Clinical + Top Env. (AUC:", round(rf_roc_selected$auc, 3), ")")),
  col = c("blue", "red", "green"),
  lty = c(1, 2, 2), lwd = 2
)

message("Full model AUC: ",                    rf_roc_full$auc)
message("Clinical Model AUC: ",                rf_roc_clinical$auc)
message("Clinical + Top Environmental AUC: ",  rf_roc_selected$auc)
```

## Variable importance

```{r rf-importance}
importance_df <- data.frame(
  Feature    = names(rf_model$variable.importance),
  Importance = as.numeric(rf_model$variable.importance)
) %>% arrange(desc(Importance))
head(importance_df, 20)

importance_df_clinical <- data.frame(
  Feature    = names(rf_model_clinical$variable.importance),
  Importance = as.numeric(rf_model_clinical$variable.importance)
) %>% arrange(desc(Importance))
importance_df_clinical

importance_df_selected <- data.frame(
  Feature    = names(rf_model_selected$variable.importance),
  Importance = as.numeric(rf_model_selected$variable.importance)
) %>% arrange(desc(Importance))
importance_df_selected
```

## DeLong tests

```{r rf-delong-tests}
# Clinical vs Full
roc_compare_1 <- roc.test(rf_roc_clinical, rf_roc_full, method = "delong")
roc_compare_1

# Clinical vs Clinical + Top environmental
roc_compare_2 <- roc.test(rf_roc_clinical, rf_roc_selected, method = "delong")
roc_compare_2
```

---

# Session info

```{r session-info}
sessionInfo()
```
